% FRAME Paper 2 — applications paper (bias-detection tool reliability + triage protocol)
% Shares corpus/models/protocol with the companion paper (p1_disclosure.tex).
\documentclass{article}
\usepackage{iclr2026_conference,times}
\usepackage{natbib}
\usepackage{amsmath,amssymb}
\usepackage{booktabs}
\usepackage{xcolor}
\usepackage{hyperref}
\usepackage{microtype}
\newcommand{\pending}[1]{\textcolor{orange!80!black}{[#1]}}
\newcommand{\kap}{\kappa}

\title{Which Bias Flags Can You Trust?\\
A Label-Free Triage Protocol for\\
Frontier-Model News-Bias Detection}
\author{Anonymous Authors\\
\texttt{anonymous@example.com}}

\begin{document}
\maketitle

\begin{abstract}
News-bias products---bias checkers, balanced-news feeds, media-literacy
tools---are increasingly built by asking a frontier model to flag biased
spans, label an article's lean, and explain itself. Which of those
outputs can a builder trust, and how should the untrustworthy ones be
triaged \emph{without} ground-truth labels at inference time? Evaluating
two frontier target models and two cross-family judges on 200
expert-rated U.S.\ news articles, we first map the primitives: the
\emph{volume} of flagged spans is a valid partisan-intensity meter
(Spearman $\rho \approx 0.6$), and models rank-track the continuous
expert lean rating at $\rho \approx 0.8$, but a single model's lean
label near the center-right, its omission/framing flags (chance-corrected
agreement $\kap \le 0.14$), and the specific text it highlights
(cross-model overlap $0.26$) are model-specific opinion. We then turn the
map into a validated selective-prediction protocol. A cost-free
\emph{committee-disagreement} bit ranks errors as well as a calibrated
confidence score globally (AUC $0.69$ vs $0.70$, $p{=}.90$) yet captures
significantly more errors per unit of human-review budget
($+6$--$8$pts, $p{<}.05$, at a $20$--$30\%$ budget), because
disagreements are far more error-dense ($32\%$ vs $71\%$ accurate); a
$4$-model unanimity gate clears $44\%$ of articles at $87\%$ accuracy;
and neither raw confidence (both models overconfident; a stated $0.9$
means ${\sim}75\%$) nor a majority-vote ensemble (no accuracy gain over
the best single model) should be used as a probability. Every result
holds across five prompt conditions. Findings are exploratory and
inherit AllSides outlet labels as ground truth; all statistics carry
bootstrap CIs.
\end{abstract}

\section{Introduction}
A growing class of products puts a frontier language model in the role
of a news-bias analyst: it highlights loaded spans, assigns a
left/right lean, summarizes ``both sides,'' and narrates why. The
implicit assumption is that these outputs are measurements. Many are
not---they are one model's opinion, and a different frontier model,
shown the same article, would say something else. A builder needs two
things before shipping: to know \emph{which} outputs reproduce, and---for
the ones that do not---a way to catch the likely errors that needs
\emph{no} ground-truth labels at inference time, since a deployed
product has none.

This paper supplies both. We do not ask whether the models are biased (a
companion paper takes up the behavioral question); we ask which of a
bias detector's outputs a tool can present as reliable, and how to
triage the rest. Our answer is a reliability map with a clean split---
primitives anchored to explicit text reproduce across independent
models; primitives that require inference about what is absent or
implied do not---and, built on it, a label-free triage protocol whose
central signal is not model confidence but \emph{model disagreement}. We
validate the protocol against the standard confidence baseline with an
operating curve, and show it is stable across prompt conditions rather
than an artifact of one.

\textbf{Contributions.}
\begin{enumerate}
  \item \textbf{A reliability map of bias-detection primitives.} Which
  outputs of a frontier-model bias detector reproduce across models
  (detection volume; continuous-rating rank; cross-model agreement) and
  which are model-specific opinion (lean labels near center-right;
  omission and framing flags; specific highlights)---each with a number
  and a bootstrap CI.
  \item \textbf{A validated, label-free triage protocol.} A cost-free
  committee-disagreement router with a published risk--coverage
  operating curve, benchmarked against a calibrated confidence router
  (equal global ranking, superior in-budget error capture); a
  high-precision unanimity gate; and a per-model calibration analysis
  showing raw confidence must be rescaled and is trustworthy only for
  direction. The protocol needs no labels at inference and is
  prompt-invariant.
  \item \textbf{A directional failure mode shared across vendors.} All
  four models read moderate-right content as center-or-left, so a
  lean-labeling tool built on them under-counts the right; stable in
  $5/5$ prompt conditions.
\end{enumerate}

\section{Setup}
We reuse the corpus, models, and rollouts of the companion study; we
summarize here and defer the full protocol (including the reliability
gates behind the judge instruments) to it. The corpus is $N{=}200$
contemporary U.S.\ news articles on a balanced $5\times8$ grid: five
AllSides-anchored lean classes (Left, Lean~Left, Center, Lean~Right,
Right; 40 each) $\times$ eight topics, each also carrying a continuous
expert lean rating; models receive article text only (no outlet or
headline). Two frontier \emph{target} models (Claude Sonnet~4.5,
GPT-4.1) and two cross-family \emph{judge} models (Claude Sonnet~4.6,
GPT-5) perform span-level bias detection with a shared bias-type schema
\citep{spinde2021babe,hamborg2019automated}, summarization, and
five-class lean classification with confidence. All analyses are
exploratory (not pre-registered) and use expert AllSides labels as
ground truth \citep{baly2020detect}; bootstrap CIs use $10^4$ resamples.

\section{Which primitives reproduce}

\textbf{Detection volume is a valid intensity meter.} The simplest
signal---how many spans a model flags---tracks how loaded an article is:
flagged-span count rises monotonically with the article's expert-rated
intensity (Spearman $\rho{=}0.64$ Sonnet, $0.58$ GPT-4.1 against
$|\text{lean rating}|$, both $p{<}10^{-18}$; partisan articles draw
$2.3\times$ the flags of Center articles). A builder can trust a flag
\emph{count} as a meter even where the individual flags are unreliable.

\textbf{Lean predictions track the continuous rating, not just the
bucket.} Each model's five-class prediction rank-correlates with the
\emph{continuous} expert rating at least as strongly as with the discrete
label (Spearman $0.77$--$0.86$ across the four models; difference vs the
label $\le 0.01$), and within a single label class the models resolve
graded intensity the bucket discards (within-Lean-Left
$\rho(\text{pred},\text{rating}){=}0.40$--$0.50$, $p\le.01$, all four).
The lean signal is thus finer than the five-way label, and the results
below are not an artifact of five-way bucketing. (We do not claim
independence from the AllSides labels: the continuous rating and the
discrete label partition the corpus identically here.)

\textbf{Only concrete, word-anchored bias forms reproduce.} Given both
target models the same bias-type schema, we measure how often they agree
a form is present (Cohen's $\kap$ over 256 jointly-rated articles). No
form clears moderate agreement---the ceiling is opinion-stated-as-fact at
$\kap{=}0.45$---and the forms defined by what is \emph{absent from} or
\emph{unverified against} the text collapse toward chance
(Table~\ref{tab:biastype}). The disagreement is not random: the two
models have distinct habits---Claude leads with bias-by-omission ($14\%$
of its flags), GPT-4.1 with sensationalism ($15\%$)---so one hunts what
a story omits while the other hunts its tone, which is exactly why they
cannot agree omission is present. The unreliability reaches the span
level: even when both flag the same article, the specific highlighted
text overlaps at only Jaccard $0.26$. For a tool: show concrete
word-level flags; treat any single model's omission, framing, or
unsupported-claim flag---and any individual highlight---as one model's
opinion, surfaced only on cross-model agreement.

\begin{table}[h]\centering\small
\begin{tabular}{lc@{\hskip 2em}lc}
\toprule
Reliably agreed ($\kap$) & & Barely above chance ($\kap$) & \\
\midrule
opinion stated as fact & $0.45$ & framing & $0.22$ \\
negativity & $0.32$ & omission of attribution & $0.19$ \\
word choice & $0.31$ & bias by omission & $0.14$ \\
sensationalism & $0.24$ & unsubstantiated claims & $0.09$ \\
\bottomrule
\end{tabular}
\caption{Cross-model agreement on bias forms (Cohen's $\kap$, two target
models, 256 shared articles). Detectability tracks textual
explicitness.}
\label{tab:biastype}
\end{table}

\section{A label-free triage protocol}

The map says most single-model outputs cannot be shown as fact. A
deployed product cannot fix this with ground truth---it has none at
inference---so it needs a trust signal computable from the model outputs
alone. We show the best such signal is \emph{model disagreement}, and
benchmark it against the obvious alternative, model confidence.

\textbf{Route disagreement first.} On five-class lean, when the two
target models agree they are $71.4\%$ accurate ($n{=}182$); when they
disagree, only $31.6\%$ ($n{=}117$; gap $39.8$pts, $95\%$ CI $[29.0,
50.4]$). As an error ranker, the free disagreement bit is
\emph{statistically indistinguishable} from a calibrated confidence
score in global discrimination (AUC $0.692$ $[.639,.744]$ vs $0.697$
$[.636,.754]$; paired difference $-0.004$, $p{=}.90$), yet it captures
more errors per unit of human-review budget in the operational regime,
because disagreements are far more error-dense than low-confidence items
(Table~\ref{tab:router}): at a $20$--$30\%$ review budget, routing
disagreement first captures $+6$ to $+8$ points more of all errors than
a confidence threshold ($p{=}.042$ and $p{=}.030$). The advantage is a
directional, in-budget claim---it narrows and loses significance once the
budget exceeds the disagreement rate, and part of the raw agree/disagree
accuracy gap is mechanical (on a disagreement, only one of two labels can
match a single ground-truth label). The operational recommendation is
nonetheless clean: \emph{route model disagreements to human review
first; confidence adds no ranking value until the budget exceeds the
disagreement rate.}

\begin{table}[h]\centering\small
\begin{tabular}{lccccc}
\toprule
Review budget & $10\%$ & $20\%$ & $30\%$ & $40\%$ & $50\%$ \\
\midrule
route by disagreement & $15.5$ & $\mathbf{31.1}$ & $\mathbf{46.6}$ & $61.3$ & $67.7$ \\
route by confidence & $15.2$ & $25.0$ & $38.6$ & $56.1$ & $65.2$ \\
\bottomrule
\end{tabular}
\caption{Percent of all classification errors captured at each
human-review budget (route least-reliable first; $n{=}299$, two target
models). Disagreement dominates confidence in the practical
$20$--$30\%$ regime (bold, $p{<}.05$).}
\label{tab:router}
\end{table}

\textbf{At committee scale, trust unanimity---not a vote.} Adding the two
judges, a four-model majority vote does \emph{not} beat the best single
model ($0.646$ vs $0.662$ exact; McNemar $p{=}.68$; bootstrap gain
$95\%$ CI $[-.061,+.030]$), and a builder cannot pre-select which model
will be best (bootstrap $P(\text{is best})$: GPT-4.1 $.63$, Sonnet~4.6
$.35$, GPT-5 $.02$, Sonnet~4.5 $.00$). Ensembling therefore buys
robustness, not accuracy. Its deployable value is instead a
high-precision \emph{gate}: the four models agree unanimously on $44\%$
of articles, and there they are $87\%$ accurate---so unanimity auto-clears
nearly half the volume at high precision, and the residual is exactly the
disagreement set the router sends to review.

\textbf{Never read raw confidence as a probability.} Both models are
severely overconfident on the exact five-class label (mean confidence
$0.85$--$0.89$ vs accuracy $0.53$--$0.63$); a stated $0.9$ empirically
means $71\%$ (Claude) to $77\%$ (GPT-4.1) exact accuracy. GPT-4.1's
confidence is the better gate (Brier $0.286$ vs $0.336$; ECE $0.260$ vs
$0.324$, the exact-target gap suggestive at $p{=}.05$), and---critically---
the overconfidence is almost entirely an \emph{exact-boundary} effect:
scored on lean \emph{direction} (within one class), GPT-4.1's confidence
is essentially calibrated (ECE $0.024$) and Claude's nearly so ($0.085$).
Confidence is a usable ordering for \emph{direction}, never a usable
probability for the exact label---rescale before thresholding.

\section{The directional failure mode}

The reproducible signals above are politically even-handed on average;
the classification \emph{errors} are not. Breaking accuracy down by true
class exposes a single hard case shared by every model
(Table~\ref{tab:byclass}): \emph{Lean~Right} is the hardest label for all
four (exact accuracy $0.07$--$0.46$), and difficulty falls as an
article's expert lean strengthens (Spearman $\rho{=}-0.37$,
$p{<}10^{-7}$)---the models read the poles well and lose the moderate
right. The misreads are \emph{directional}: misclassified Lean-Right
articles are pulled \emph{left}, not merely blurred to center---$33\%$
are labeled Lean-Left or Left and $21\%$ Center, while Center articles
are called Lean-Left $32\%$ of the time versus Lean-Right $4\%$. A
``balanced feed'' or lean-labeling product built on these
classifications therefore \emph{systematically under-counts right-leaning
content}, consistent with concurrent reports of ``center-collapse''
\citep{kennedy2026left}, which our per-class decomposition refines into a
right-to-left asymmetry.

\begin{table}[h]\centering\small
\begin{tabular}{lccccc}
\toprule
Model & Left & Lean~L & Center & \textbf{Lean~R} & Right \\
\midrule
Claude Sonnet 4.5 & $.97$ & $.23$ & $.50$ & $\mathbf{.07}$ & $.90$ \\
GPT-4.1 & $.88$ & $.47$ & $.75$ & $\mathbf{.46}$ & $.72$ \\
Claude Sonnet 4.6 & $.92$ & $.78$ & $.55$ & $\mathbf{.20}$ & $.80$ \\
GPT-5 & $.82$ & $.70$ & $.53$ & $\mathbf{.42}$ & $.60$ \\
\bottomrule
\end{tabular}
\caption{Exact classification accuracy by true lean class. Lean~Right is
the hardest label for every model.}
\label{tab:byclass}
\end{table}

\textbf{Both the map and the failure are prompt-invariant.} Re-running
the two headline results under five different prompt scaffolds, the
agreement$\to$accuracy gap holds in all five (gap $0.28$--$0.36$,
bootstrap $P(\text{gap}{>}0){=}1.0$ in every condition) and Lean~Right is
the strict hardest class in all five (pooled accuracy $0.21$--$0.29$).
The pattern is a property of the models, not of one prompt. Two honest
riders: the Lean-Right effect is Claude-driven (Claude is the worst model
on it in $5/5$ conditions, GPT-4.1 in only $3/5$), and lean
classification is least reliable on policy-abstract topics
(elections/economy/climate) and most reliable on identity-concrete ones.

\section{Discussion and Limitations}
These results are a reliability map and a triage protocol, not a bias
verdict: they tell a builder which parts of a frontier-model bias
detector today's models can power, and---for the rest---how to catch
likely errors with a signal (\emph{model disagreement}) that costs
nothing and needs no labels at inference. The protocol's core claims are
selective-prediction claims validated against the standard confidence
baseline with an operating curve, and shown prompt-invariant; that
methodology, and the balanced expert-rated corpus, are the durable
contribution, while the specific per-model numbers are a dated snapshot.

Three limits bound the claims. All findings are \emph{exploratory} and
inherit AllSides' outlet-anchored labels as the sole ground truth
\citep{baly2020detect}; the directional error could partly reflect
article-vs-outlet label drift, though uniform label noise cannot produce
a one-sided right-to-left pattern, and a second independent label source
(e.g.\ per-article human consensus) is the highest-value next test. The
router is validated on the baseline condition and one model pair; and the
label-free signals so far require a \emph{committee}---disagreement needs
two models, the unanimity gate four. A genuine single-model cold-start
signal (self-consistency across temperature-sampled generations,
benchmarked against the disagreement router's AUC of $0.69$) is the
natural next step. Finally, any LLM judge inside a tool's own
quality-control pipeline inherits the reliability problem we document for
the models themselves, and needs the pre-committed validation the
companion paper specifies.

\bibliography{references}
\bibliographystyle{iclr2026_conference}
\end{document}
